Pool Model Paper

### 2023-08-31 pool_model.tex
spatial limitation: correct mistake in ODEs
3 pool model, 1 species corrections

\section{General Mathemetical Formulation}
Consider the system conssisting of $N$ species in $M$ different states.
In the general representation, the system can be described with the Ordinary Differential Equations (ODEs) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
   \label{eq:model_ODE}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector descibing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_M  \\
        \vdots &       & \vdots \\
        x^N_1  & \dots & x^N_M  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{ij}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentartion of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu, ...).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.

@@

To calculate the bacterial count for one species $i$, we simply need to sum the tensor/vector $\mbx$ over all possible states:
\begin{equation}
    N_n = \sum_m \mbx_i^n.
\end{equation}
While the total bacterial count of all species of our system is
\begin{equation}
    N_t = \sum_j \sum_m \mbx_i^j.
\end{equation}
In the following section we demonstate the application of this approach for different toy models describing the processes within and between bacteria communities.

@@

Consider for simplicity the system where only one bacteria species is present $N=1$.
The bacterial growth curves usually consists of different consecutive phases~\cite{buchanan_when_1997}.
At the beginning no increase in bacterial count is present and so-called lag-phase is observed.
This phenomenon appears due to adaptation of the microbial community, for example, to the new environmental conditions.
This so-called preparation to the exponential growth includes synthesys of the needed compounds, initiation of the transciption~\cite{rolfe_lag_2012} and poorly studied physiological and regulatory mechanisms~\cite{monod_growth_1949,}
\cmt{For example, in Baranyi and Roberts modelling of lag-phase ???(maybe add).}
After this, the succeeded in adaptation bacteria enter the exponential growth phase, where the process of cell division is directly happening \cmt{Add Ref}.
When the nutrition is exhausted by the growing bacteria, the stationary state of the growth curve occurres.
In this state the cultures accumulate waste products of metabolitic activity that leads to cell death of the bacteria.
At the same time dying cells release the nutritions for survivors to continue growing.
As a result the balance between the growing and dyuing processes leads to the constant with time bacterial count~\cite{navarro_llorens_stationary_2010}.
In the described simplified version of the real-life system three possible bacteria states are mentioned corresponding to which process is happeng: adaplation, growth or death.
Motivated by this we divide a bacterial population into three pools (states) (\textbf{Figure \ref{fig:SchematicRep}}): 
\begin{itemize}
\item $L(t)$: the fraction of the population in the lag phase
\item $G(t)$: the fraction of the population growing and dividing
\item $D(t)$: the fraction of the population undergoing cell death
\end{itemize}
Here the concentrations of the pools are components of our variable vector $\mbx (t) =\{L(t), G(t), D(t)\}$.

\begin{figure}[t]
    \begin{center}
    \includegraphics[width=0.7\textwidth]{Figures/TPM_fig.pdf}
    \caption{Schematic representation of exchange between three states or pools.}
    \label{fig:SchematicRep}
    \end{center}
\end{figure}

To detremine the function $\mbF(\mbx, \mbm, \mbp, \mbu)$ consider the flows between different states.
The pools $G$ and $D$ are unidirectionally connected with the pool $L$. 
For now we assume for simplicity that there is no back flow between growth state $G$ and lag-state $L$. 
The suggested model can be cast in the following reactions:
\begin{eqnarray}
L &\stackrel{\lambda}{\longrightarrow} & G\\
L &\stackrel{\mu}{\longrightarrow} & D\\
G &\stackrel{\alpha}{\longrightarrow} & 2G\\
G &\stackrel{\mu'}{\longrightarrow} & D\\
D &\stackrel{\beta}{\longrightarrow} & \emptyset
\end{eqnarray}.

Then the ODEs describing such a model can be written as 
\begin{equation}
\begin{cases}
    \dot{L} = -(\mu + \lambda) L\\
    \dot{G} = \lambda L + \alpha G - \mu' G\\
    \dot{D} = \mu  L + \mu' G- \beta D  
\end{cases}.
\end{equation}

\cmt{Add plots ?? Or already after resource is added?}

\subsubsection{Resource Dependence}

Based on experimental observations, it was shown that the bacterial growth is strongly influenced by the limiting resource.
In the system where all needed nutritients are in excess except for one the total growth is linear to the conentration of this liminting nutritient~\cite{hibbing_bacterial_2010, monod_growth_1949}.
The control of the population density with respect to the available nutritions is implemented using cell communication process called quorum sensing. 
Using release of the extracellular signalling molecules (autoinducers) it is possible to synchronize the gene expression of the whole group~\cite{ng_bacterial_2009}.

@@

Note that the pools $L$, $G$ and $D$ represent number of bacteria (or number of bacteria per unit area/volume) while $R_0$ represents an abstract resource pool. 
Let say $G=N_B/V$ and $[G]=1/V$, where $N_B$ is number of bacteria and $V$ is the volume. 
Then $R_0=N_R/V$ and $[R_0]=1/V$, where $N_R$ is the number of resource molecules (or any appropriate unit, e.g., mol). 
Then $[\alpha_0]=V/s$, $[N_t]=1/V$, and $[\alpha]=1/s$. 
For homogeneous densities it is equivalent to consider the total amount, i.e., multiplying $L$ and $G$ by the volume. 
In this case $[L]=[G]=[N_t]=1$.
For simplicity we rather make a transition to the dimensionless constant $\alpha:=\alpha_0 N_t$ and diminsioneless resource pool $R(t) = R_0 / N_t$ ($0 \leqslant R \leqslant 1$).
Here the full capacity of the resource pool $N_t$ was used as a scaling constant.
Then the growth term can be simply written as $\alpha G R$, and the microenvironment of the system introduced by \ref{eq:model_microenv} is defined by ODE:
\begin{equation}
\dot{R} = -\frac{\alpha}{N_t} R G
\end{equation}

Explointing that limitation of the resource pool also limits the total becterial count of the system  we find:
\begin{equation}
    R = 1 - \frac{L+G+D}{N_t}
\end{equation}
And, hence, the full system can be written as
\begin{equation}
    \begin{cases}
        \dot{L} = -(\mu + \lambda) L\\
        \dot{G} = \lambda L + \alpha G\left(1-\frac{L+G+D}{N_t}\right)-\mu' G\\
        \dot{D} = \mu  L + \mu' G- \beta D  
    \end{cases}.
\end{equation}

\subsubsection{Lag-time Estimation (??)}

Note that one could further simplify this model to a two pool or state model by omitting the pool $D$ and removing cells with rate $\beta$ directly from pool $L$. 
The relevant question here is whether cells in the pool $D$ still contribute to the limitation of the resources by, e.g., occupying space or consuming important nutrients. 
If yes, the three pool model would be advantageous, if not the two pool model is sufficient with the advantages of having one parameter less.\\

The dynamics of the three pool model can be captures by the coupled system of ordinary differential equations (the two pool model can be achieved by setting $D\equiv 0$ and replacing $\mu$ with $\beta$):
to be solved with the initial conditions $L(0)=n_0$, $G(0)=0$, and $D(0)=0$, where $n_0$ is the initial bacterial population. 
$N_t$ denotes the maximal size of the total bacterial population due to environmental limitations (\textbf{Figure \ref{GrowthDip}}). 
For $\mu'=0$ the equations for $L$ and $D$ can be readily integrated resulting in:
\begin{eqnarray}
L(t) &=& n_0 e^{-(\mu+\lambda)t}\\
D(t) &=& n_0 \mu e^{-\beta t}\frac{1-e^{-(\mu+\lambda-\beta)t}}{\mu+\lambda-\beta}.
\end{eqnarray}
For constant parameters $\mu$, $\lambda$, and $\beta$ the integrals for $L$ and $D$ can be put into the differential equation for $G$. 
However, for time dependent parameters, e.g. for a dependence on a non-static, dynamic temperature, it appears to be not advantageous to use the analytical results for $L$ and $D$ and work instead with all three equations for $L$, $G$, and $D$. 
It is interesting to see what type of equations result for the total bacterial population $n=L+G+D$:
\begin{eqnarray}
\dot{G} &=& \lambda L + \alpha G\left(1-\frac{n}{N_t}\right)\\
\dot{n} &=& - \beta D + \alpha G\left(1-\frac{n}{N_t}\right). 
\end{eqnarray}
with the initial conditions $G(0)=0$ and $n(0)=n_0$. 
These equations have a clear interpretation, are derived from simple principles, and are different from the popular Barayni-Roberts and modified-Gompertz models. 
A notable difference is that the growth rate $\alpha$ is time independent. 
For the functions $L$ and $D$ it holds: $\lim_{t\to\infty} L(t) = \lim_{t\to\infty} D(t) = 0$. 
It follows immediately that $\lim_{t\to\infty} G(t) = N_t$. 
The equation exhibit a lag phase and an initial drop due to cell death. 
To find a substantial initial drop one needs $\beta \mu \gg \alpha \lambda$. 
The lag time $t_L$ can be approximated by the time at which half of the initial population is in the growth phase, i.e., $n_0=2G(t_L)$. 
Assuming $L\ll N_t$ we can ignore the non-linear term in the equation for $G$, which gives rise to:
\begin{eqnarray}
1&=& 2\int_0^{t_L} \lambda L(t)e^{\alpha(t_L-t)}dt\\
1  &=& 2\lambda e^{\alpha t_L}\frac{1-e^{-(\mu+\lambda+\alpha)t_L}}{\mu+\lambda+\alpha}.
\end{eqnarray}
For $\alpha > \mu+\lambda$ $t_L$ can be approximated by:
\begin{eqnarray}
t_L &\approx& -\frac{1}{\alpha}\ln\left(\frac{2\lambda}{\mu+\lambda+\alpha}\right).
\end{eqnarray}
The lag time $t_L$ diverges very slowly, logarithmically with $\lambda\to 0$.

\cmt{Add/make plots to the model 3 pool/ 2 pool.}

\cmt{Add limitation due to resource}

@@

Another way to encorporate the resource dependence is to assume that the bacteria only leave the lag-phase if there are nutrients, i.e., resources. 
Otherwise they stay in the lag-phase.
Then using the model consisted of two bacteria states $L$ and $G$ we can define the system as
\begin{eqnarray}
\dot{L} &=& - \lambda R L\\
\dot{G} &=& \lambda R L +\alpha R G\\
\dot{R} &=&-\frac{\alpha}{N_t} R G.
\end{eqnarray}

Turning back to the equations above we can state: $L+G+N_tR=N_t$. For $t=0$ we have $n_0+N_tR_0=N_t$ (given $L_0=n_0$ and $G_0=0$) and from this: $N_tR_0=N_t-n_0$. Using this we eliminate $R$ and arrive at: 
\begin{eqnarray}
\dot{L} &=& - \lambda \left(1-\frac{L+G}{N_t}\right)L\\
\dot{G} &=& \lambda \left(1-\frac{L+G}{N_t}\right)L +\alpha \left(1-\frac{L+G}{N_t}\right) G
\end{eqnarray}
The steady state is $L_s+G_s=N_t$. However, it depends on the initial conditions and the parameters what $L_s$ and $G_s$ are. There is no way to determine $L_s$ and $G_s$ besides solving the dynamical equations, i.e., the steady state cannot be determined directly by solving a set of algebraic equations. To see this we determine the nullclines: the first nullcline is $G=N_t-L$ and the second is $G=N_t-L$, i.e., the nullclines coincides. Every point on the line $G=N_t-L$ is a fixed point. 
This can be seen in Fig. \ref{Phaseplot}. 
For $\lambda\ll \alpha$ the lag-pool is not emptied before all resources are consumed. As a consequence the lag-pool is non-zero in steady-state. 
In contrast, for $\lambda\gg \alpha$ the lag-pool is depleted before the resources are consumed, i.e., the lag-pool is zero in steady state.

@@

\subsection{Back flow from G to L}

@@

The ordinary differential equations for the pool model with environmental-shock based back-flow read:
\begin{eqnarray}
\dot{L} &=&\gamma(t) G - (\mu + \lambda) L\\
\dot{G} &=& \lambda L -\gamma(t) G + \alpha G\left(1-\frac{L+G+D}{N_t}\right)\\
\dot{D} &=& \mu  L - \beta D
\end{eqnarray}

@@

However, except the inhibition due to resource competition, bacteria developed differnt others active competition mechanisms.
One of the examples of such mechanisms is a production of the antimicrobial compounds, which can be also called antibiotics or inhibitors.

@@

\cmt{\cite{lenski_coexistence_1986}: Modelling of the 2 sp. system with 1resource and 1 inhibitor.}

Now extend our model where bacteria not only consume nutritions from resource pool $R$ but one of the species also actively produce the inhibiting compounds $I$. 
It is the accumulation of the bacteria waste which is responsible for the spoilage of food products. 
In order to take this into account we add an equation for the waste production:
\begin{eqnarray}
\dot{L} &=& - \lambda_f R L + \mathcal{B}(R)G\\
\dot{G} &=& \lambda_f R L +\alpha R G-\mathcal{B}(R)G\\
\dot{R} &=&-\frac{\alpha}{N_t} R G\\
\dot{I} &=& \mathcal{F}(R,G).
\end{eqnarray}
The function $\mathcal{F}$ denotes the model for the antibiotic production rate. 
We shall discuss two possible models:
\begin{eqnarray}
\mathcal{F}_1(R,G)&=&\kappa\alpha R G=-\kappa N_t\dot{R}\\
\mathcal{F}_2(R,G)&=&\kappa G
\end{eqnarray}
Model $\mathcal{F}_1$ results in waste accumulation in steady state $I_1^s=\lim_{t\to\infty}I_1(t)=\kappa(N_t-n_0)$. 
To see this we note $\dot{I}_1=\kappa(\dot{L}+\dot{G})$, which yields: $I_1(t)=\kappa(L(t)+G(t)-n_0)$ where we used $I_1(0)=G(0)=0$, $L(0)=n_0$. 
This model assumes that inhibitor is only produced when the bacteria grow and divide. \\

Model $\mathcal{F}_2$ results in $I_2(t)=\kappa\int_{t_0}^tG(t')dt'$. 
As long as $G>0$ inhibitor is produced, no matter whether the amount of bacteria is still growing. 
The amount of inhibitor depends therefore on the back-flow model employed, which can be seen in Fig. \ref{Waste}.

\cmt{Look is this inhibition the same as waste production. Or change it.}

>>>>

\section{General Mathemetical Formulation}
Consider the system conssisting of $N$ species in $M$ different states.
In the general representation, the system can be described with the Ordinary Differential Equations (ODEs) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
   \label{eq:model_ODE}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector descibing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentartion of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.

@@

To calculate the bacterial count for one species $i$, we simply need to sum the tensor/vector $\mbx$ over all possible states:
\begin{equation}
    N_i = \sum_k \mbx_i^k.
\end{equation}
While the total bacterial count of all species of our system is
\begin{equation}
    N = \sum_i \sum_k \mbx_i^k.
\end{equation}
In the following section we demonstate the application of this approach for different toy models describing the processes within and between bacteria communities.

@@

Consider for simplicity the system where only one bacteria species is present $N=1$.
The bacterial growth curves usually consists of different consecutive phases~\cite{buchanan_when_1997}.
At the beginning no increase in bacterial count is present and so-called lag-phase is observed.
This phenomenon appears due to adaptation of the microbial community, for example, to the new environmental conditions.
This so-called preparation to the exponential growth includes synthesys of the needed compounds, initiation of the transciption~\cite{rolfe_lag_2012} and poorly studied physiological and regulatory mechanisms~\cite{monod_growth_1949,}
\cmt{For example, in Baranyi and Roberts modelling of lag-phase ???(maybe add).}
After this, the succeeded in adaptation bacteria enter the exponential growth phase, where the process of cell division is directly happening \cmt{Add Ref}.
When the nutrition is exhausted by the growing bacteria, the stationary state of the growth curve occurres.
In this state the cultures accumulate waste products of metabolitic activity that leads to cell death of the bacteria.
At the same time dying cells release the nutritions for survivors to continue growing.
As a result the balance between the growing and dyuing processes leads to the constant with time bacterial count~\cite{navarro_llorens_stationary_2010}.
In the described simplified version of the real-life system three possible bacteria states are mentioned corresponding to which process is happeng: adaplation, growth or death.
Motivated by this we divide a bacterial population into three pools (states) (\textbf{Figure \ref{fig:SchematicRep}}): 
\begin{itemize}
\item $x^1_1 (t) = L(t)$: the fraction of the population in the lag phase
\item $x^2_1 (t) = G(t)$: the fraction of the population growing and dividing
\item $x^3_1 (t) = D(t)$: the fraction of the population undergoing cell death
\end{itemize}
Here the concentrations of the pools are components of our variable vector $\mbx (t) = (L(t), G(t), D(t))^T$.

\begin{figure}[t]
    \begin{center}
    \includegraphics[width=0.7\textwidth]{Figures/TPM_fig.pdf}
    \caption{{\footnotesize Schematic representation of exchange between three states or pools.}}
    \label{fig:SchematicRep}
    \end{center}
\end{figure}

To detremine the function $\mbF(\mbx, \mbm, \mbp, \mbu)$ consider the flows between different states.
The pools $G$ and $D$ are unidirectionally connected with the pool $L$. 
For now we assume for simplicity that there is no back flow between growth state $G$ and lag-state $L$. 
The suggested model can be cast in the following reactions:
\begin{eqnarray}
L &\stackrel{\lambda}{\longrightarrow} & G\\
L &\stackrel{\mu}{\longrightarrow} & D\\
G &\stackrel{\alpha}{\longrightarrow} & 2G\\
G &\stackrel{\mu'}{\longrightarrow} & D\\
D &\stackrel{\beta}{\longrightarrow} & \emptyset
\end{eqnarray}

Then the ODEs describing such a model can be written as 
\begin{equation}
\begin{cases}
    \dot{L} = -(\mu + \lambda) L\\
    \dot{G} = \lambda L + \alpha G - \mu' G\\
    \dot{D} = \mu  L + \mu' G- \beta D  
\end{cases}
\end{equation}
or in matrix form it shows the linear dependency
\begin{equation}
    \dot \mbx  = \begin{pmatrix}
                    -(\mu + \lambda) & 0             & 0      \\
                    \lambda          & \alpha - \mu' & 0      \\
                    \mu              & \mu'          & -\beta 
                 \end{pmatrix} \mbx .
\end{equation}

\subsection{Resource Dependence}

Based on experimental observations, it was shown that the bacterial growth is strongly influenced by the limiting resource.
In the system where all needed nutritients are in excess except for one the total growth is linear to the concentration of this liminting nutritient~\cite{hibbing_bacterial_2010, monod_growth_1949}.
The control of the population density with respect to the available nutritions is implemented using cell communication process called quorum sensing. 
Using release of the extracellular signalling molecules (autoinducers) it is possible to synchronize the gene expression of the whole group~\cite{ng_bacterial_2009}.

@@

Note that the pools $L$, $G$ and $D$ represent number of bacteria (or number of bacteria per unit area/volume) while $R_0$ represents an abstract resource pool. 
Let say $G=N_B/V$ and $[G]=1/V$, where $N_B$ is number of bacteria and $V$ is the volume. 
Then $R_0=N_R/V$ and $[R_0]=1/V$, where $N_R$ is the number of resource molecules (or any appropriate unit, e.g., mol). 
Then $[\alpha_0]=V/s$, $[N_t]=1/V$, and $[\alpha]=1/s$. 
For homogeneous densities it is equivalent to consider the total amount, i.e., multiplying $L$ and $G$ by the volume. 
In this case $[L]=[G]=[N_t]=1$.
For simplicity we rather make a transition to the dimensionless constant $\alpha:=\alpha_0 N_t$ and diminsioneless resource pool $R(t) = R_0 / N_t$ ($0 \leqslant R \leqslant 1$).
Here the full capacity of the resource pool $N_t$ was used as a scaling constant.
In this case the growth term can be simply written as $\alpha G R$ and the model is fully formulated in the following way: 
\begin{eqnarray}
    \dot \mbx  &=& \mbF(\mbx, \mbm, \mbp, \mbu) = \begin{pmatrix}
                                                    -(\mu + \lambda) & 0               & 0      \\
                                                    \lambda          & \alpha R - \mu' & 0      \\
                                                    \mu              & \mu'            & -\beta 
                                                \end{pmatrix} \mbx  \\
    \dot \mbm &=& \mbH (\mbx, \mbm, \mbp, \mbu) = -\frac{\alpha}{N_t} R G.
\end{eqnarray}

Such model can be simplified by explointing that limitation of the resource pool also limits the total bacterial count of the system:
\begin{equation}
    R = 1 - \frac{L+G+D}{N_t}
\end{equation}
and, hence, the full system can be written as
\begin{equation}
    \begin{cases}
        \dot{L} = -(\mu + \lambda) L\\
        \dot{G} = \lambda L + \alpha G\left(1-\frac{L+G+D}{N_t}\right)-\mu' G\\
        \dot{D} = \mu  L + \mu' G- \beta D 
        \label{eq:3pool_resource} 
    \end{cases}.
\end{equation}

\begin{figure}[t]
    \begin{center}
    \includegraphics[width=0.67\textwidth]{Figures/pool_model_3pools_resource.pdf}
    \caption{{\footnotesize The time dependence of the total bacterial count $N(t) = \sum_i \sum_k \mbx_i^k = L(t) + G(t) + D(t)$ is obtained as a sum of bacteria in all states.
    All curves were calculated using the model (\ref{eq:3pool_resource}) with parameters $\alpha=2.5$, $\mu'=0.5$ and $\beta=5$.
    The purple curve shows significant dip in behaviour of the system where the death of the bacteria is present $\mu = 0.5$.
    Other curves show the case of no bacteria dying $\mu=0$ and represent different length of lag phases:
    short $\lambda=10^{-2}$ (red), intermediate $\lambda=10^{-3}$ (blue) and long one $\lambda=10^{-4}$ (green).
}}
    \label{fig:3pool_resource_plots}
    \end{center}
\end{figure}

\subsection{Lag-time Estimation}

Note that one could further simplify this model to a two pool or state model by omitting the pool $D$ and removing cells with rate $\beta$ directly from pool $L$. 
The relevant question here is whether cells in the pool $D$ still contribute to the limitation of the resources by, e.g., occupying space or consuming important nutrients. 
If yes, the three pool model would be advantageous, if not the two pool model is sufficient with the advantages of having one parameter less.

The dynamics of the three pool model can be captures by the coupled system of ordinary differential equations (the two pool model can be achieved by setting $D\equiv 0$ and replacing $\mu$ with $\beta$):
to be solved with the initial conditions $L(0)=n_0$, $G(0)=0$, and $D(0)=0$, where $n_0$ is the initial bacterial population. 
$N_t$ denotes the maximal size of the total bacterial population due to environmental limitations (\textbf{Figure \ref{fig:3pool_resource_plots}}). 
For $\mu'=0$ the equations for $L$ and $D$ can be readily integrated resulting in:
\begin{equation}
    \begin{cases}
        L(t) = n_0 e^{-(\mu+\lambda)t}\\
        D(t) = n_0 \mu e^{-\beta t}\frac{1-e^{-(\mu+\lambda-\beta)t}}{\mu+\lambda-\beta}
    \end{cases}.
\end{equation}
For constant parameters $\mu$, $\lambda$, and $\beta$ the integrals for $L$ and $D$ can be put into the differential equation for $G$. 
However, for time dependent parameters, e.g., for a dependence on a non-static, dynamic temperature, it appears to be not advantageous to use the analytical results for $L$ and $D$ and work instead with all three equations for $L$, $G$, and $D$. 
It is interesting to see what type of equations result for the total bacterial population $n=L+G+D$:
\begin{equation}
    \begin{cases}
        \dot{G} = \lambda L + \alpha G\left(1-\frac{n}{N_t}\right)\\
        \dot{n} = - \beta D + \alpha G\left(1-\frac{n}{N_t}\right)
    \end{cases}
\end{equation}
with the initial conditions $G(0)=0$ and $n(0)=n_0$. 
These equations have a clear interpretation, are derived from simple principles, and are different from the popular Barayni-Roberts and modified-Gompertz models. 
A notable difference is that the growth rate $\alpha$ is time independent. 
For the functions $L$ and $D$ it holds: $\lim_{t\to\infty} L(t) = \lim_{t\to\infty} D(t) = 0$. 
It follows immediately that $\lim_{t\to\infty} G(t) = N_t$. 
The equation exhibit a lag phase and an initial drop due to cell death. 
To find a substantial initial drop one needs $\beta \mu \gg \alpha \lambda$. 
The lag time $t_L$ can be approximated by the time at which half of the initial population is in the growth phase, i.e., $n_0=2G(t_L)$. 
Assuming $L\ll N_t$ we can ignore the non-linear term in the equation for $G$, which gives rise to:
\begin{eqnarray}
1&=& 2\int_0^{t_L} \lambda L(t)e^{\alpha(t_L-t)}dt\\
1  &=& 2\lambda e^{\alpha t_L}\frac{1-e^{-(\mu+\lambda+\alpha)t_L}}{\mu+\lambda+\alpha}.
\end{eqnarray}
For $\alpha > \mu+\lambda$ $t_L$ can be approximated by:
\begin{eqnarray}
t_L &\approx& -\frac{1}{\alpha}\ln\left(\frac{2\lambda}{\mu+\lambda+\alpha}\right).
\end{eqnarray}
The lag time $t_L$ diverges very slowly, logarithmically with $\lambda\to 0$.

@@

Another way to encorporate the resource dependence is to assume that the bacteria only leave the lag-phase if there are nutrients, i.e., resources. 
Otherwise they stay in the lag-phase.
Then using the model consisted of two bacteria states $L$ and $G$ we can define the system as
\begin{equation}
    \begin{cases}
        \dot{L} = - \lambda R L\\
        \dot{G} = \lambda R L +\alpha R G\\
        \dot{R} =-\frac{\alpha}{N_t} R G.
    \end{cases}
\end{equation}

Turning back to the equations above we can state: $L+G+N_tR=N_t$. For $t=0$ we have $n_0+N_tR_0=N_t$ (given $L_0=n_0$ and $G_0=0$) and from this: $N_tR_0=N_t-n_0$. Using this we eliminate $R$ and arrive at: 
\begin{equation}
    \begin{cases}
        \dot{L} = - \lambda \left(1-\frac{L+G}{N_t}\right)L\\
        \dot{G} = \lambda \left(1-\frac{L+G}{N_t}\right)L +\alpha \left(1-\frac{L+G}{N_t}\right) G
\end{cases}
\end{equation}
The steady state is $L_s+G_s=N_t$. However, it depends on the initial conditions and the parameters what $L_s$ and $G_s$ are. There is no way to determine $L_s$ and $G_s$ besides solving the dynamical equations, i.e., the steady state cannot be determined directly by solving a set of algebraic equations. To see this we determine the nullclines: the first nullcline is $G=N_t-L$ and the second is $G=N_t-L$, i.e., the nullclines coincides. Every point on the line $G=N_t-L$ is a fixed point. 
This can be seen in Fig. \ref{Phaseplot}. 
For $\lambda\ll \alpha$ the lag-pool is not emptied before all resources are consumed. As a consequence the lag-pool is non-zero in steady-state. 
In contrast, for $\lambda\gg \alpha$ the lag-pool is depleted before the resources are consumed, i.e., the lag-pool is zero in steady state.

@@

\subsection{Back flow from Growth to Lag pool}

@@

The ordinary differential equations for the pool model with environmental-shock based back-flow read:
\begin{equation}
    \begin{cases}
    \dot{L} =\gamma(t) G - (\mu + \lambda) L\\
    \dot{G} = \lambda L -\gamma(t) G + \alpha G\left(1-\frac{L+G+D}{N_t}\right)\\
    \dot{D} = \mu  L - \beta D
    \end{cases}
\end{equation}

@@

However, except the inhibition due to resource competition, bacteria developed differnt others active competition mechanisms.
One of the examples of such mechanisms is a production of the antimicrobial compounds, which can be also called antibiotics or inhibitors. \cmt{toxins}

@@

\subsubsection{Toxins Production}
Now extend our model where bacteria not only consume nutritions from resource pool $R$ but one of the species also actively produce the inhibiting compounds $T$. 
It is the accumulation of the bacteria waste which is responsible for the spoilage of food products. 
In order to take this into account we add an equation for the waste production:
\begin{eqnarray}
\dot{L} &=& - \lambda_f R L + \mathcal{B}(R)G\\
\dot{G} &=& \lambda_f R L +\alpha R G-\mathcal{B}(R)G\\
\dot{R} &=&-\frac{\alpha}{N_t} R G\\
\dot{T} &=& \mathcal{F}(R,G).
\end{eqnarray}
The function $\mathcal{F}$ denotes the model for the antibiotic production rate. 
We shall discuss two possible models:
\begin{eqnarray}
\mathcal{F}_1(R,G)&=&\kappa\alpha R G=-\kappa N_t\dot{R}\\
\mathcal{F}_2(R,G)&=&\kappa G
\end{eqnarray}
Model $\mathcal{F}_1$ results in waste accumulation in steady state $I_1^s=\lim_{t\to\infty}I_1(t)=\kappa(N_t-n_0)$. 
To see this we note $\dot{T}_1=\kappa(\dot{L}+\dot{G})$, which yields: $I_1(t)=\kappa(L(t)+G(t)-n_0)$ where we used $T_1(0)=G(0)=0$, $L(0)=n_0$. 
This model assumes that inhibitor is only produced when the bacteria grow and divide. \\

Model $\mathcal{F}_2$ results in $I_2(t)=\kappa\int_{t_0}^tG(t')dt'$. 
As long as $G>0$ toxin is produced, no matter whether the amount of bacteria is still growing. 
The amount of toxin depends therefore on the back-flow model employed, which can be seen in Fig. \ref{Waste}.

\subsubsection{Species Inhibition}
\cmt{\cite{lenski_coexistence_1986}: Modelling of the 2 sp. system with 1resource and 1 inhibitor.}
The simplest way to introduce inhibition of species A by species B is via a growth rate of species A depending on the abundance of species B. 
For sake of simplicity we ignore the lag pool and write for a mutual inhibition:
\begin{eqnarray}
\dot{G}_A &=& \mathcal{B}_A(G_B)G_A\left(1 - \frac{G_A+G_B}{N_t}\right )\\
\dot{G}_B &=& \mathcal{B}_B(G_A) G_B\left(1-\frac{G_A+G_B}{N_t}\right ). 
\end{eqnarray}
The function $\mathcal{B}_{A/B}$ captures the inhibiting effects in an effective way. 
It is straight forward to make the mechanism of inhibition more explicit by, e.g., introducing a waste pool. 
One possible choice for $\mathcal{B}$ is: 
\begin{eqnarray}
\mathcal{B}(G)&=&\frac{\alpha}{1+KG}. 
\end{eqnarray}
This corresponds to the case $W\sim G$, where $W$ is the waste (or whatever one chooses to call it) of species A affecting species B. 
If the waste accumulates one needs to replace $G$ by $\int G dt$. Keeping it simple yields:
\begin{eqnarray}
\dot{G}_A &=& \frac{\alpha_A G_A}{1+K_BG_B}\left(1 - \frac{G_A+G_B}{N_t}\right )\\
\dot{G}_B &=& \frac{\alpha_B G_B}{1+K_AG_A}\left(1-\frac{G_A+G_B}{N_t}\right ). 
\end{eqnarray}
One can distinguish two limiting cases: {\it i}) $K_{A/B}\ll N_t$ and {\it ii}) $K_{A/B}\gg N_t$. 
In the first case the inhibition is of minor importance and could be ignored. 
In the latter case one can simplify the equations, given $n_A$ and $n_B$ are non-zero:
\begin{eqnarray}
\label{Mut_Inhib}
\dot{G}_A &=&\tilde{\alpha}_A\frac{G_A}{G_B}\left(1 - G_A-G_B\right )\\
\dot{G}_B &=& \tilde{\alpha}_B\frac{G_B}{G_A}\left(1-G_A-G_B\right ). 
\end{eqnarray}
We introduced the rescaled growth rates $\tilde{\alpha}_A=\alpha_A/(N_tK_B)$ and $\tilde{\alpha}_B=\alpha_B/(N_tK_A)$. Note that $[\tilde{\alpha}]=\mathrm{s}^{-1}$. Rescaling time with $\tilde{\alpha}_A$ and defining the dimensionless parameter 
\begin{eqnarray}
\psi&=&\tilde{\alpha}_B/\tilde{\alpha}_A\\
\psi&=&\frac{K_B\alpha_B}{K_A\alpha_A}
\end{eqnarray}
results in: 
\begin{eqnarray}
\dot{G}_A &=&\frac{G_A}{G_B}\left(1 - G_A-G_B\right )\\
\dot{G}_B &=&\psi\frac{G_B}{G_A}\left(1-G_A-G_B\right ). 
\end{eqnarray}
We can separate three cases: 
\begin{eqnarray}
\psi&>&1:\quad\text{Species B growth faster and/or inhibits A stronger}\\
\psi&=&1:\quad\text{A and B are equivalent}\\
\psi&<&1:\quad\text{Species A growth faster and/or inhibits B stronger}
\end{eqnarray}
